Towards cross-middleware authentication and single sign-on for ARC Grid middleware

### 2009-04-11 doc/sec/papers/ISC09/ISC09.tex
Modification for last version

\section{Introduction} 
\label{sec:intro}
Grid provides the technology for wide-scale, cross-domain collaboration where users work with resources and share data which are distributed across administrative domains. This kind of collaboration is called Virtual Organization (VO)~\cite{IFoster01}. The resources that contribute to the VO can come from independent administrative domains, and even require different security approaches to access control due to usage of different underlying middlewares. The same heterogeneous situation applies to users who need to access the VO resources, because users holding one kind of a credential may need to utilize multiple resources with different security approaches. To solve the above issues, Grid middlewares should provide the following abilities:
\begin{itemize}
\item The user which participates a VO should be able to access data and resources without being required to constantly do authentication at each resource or service;
\item The VO should be able to easily manage the privileges and memberships of the users;
\item The VO together with its resources should be able to easily manage the authorization, given the condition that the resources are owned by different administrative domains.
\end{itemize}
In the current Grid community, the Grid Security Infrastructure (GSI)~\cite{IFoster98} is the de facto standard for authentication and transport level security. It builds on the Public Key Infrastructure (PKI), an architecture based on the X.509 public key certificate. GSI implements some enhancement (such as certificate delegation) based on the standard SSL/TLS protocol. Mutual authentication is required by GSI, and is the default configuration for GSI based Grid deployment. The X.509 certificates are required for both the client and service sides in order to achieve mutual authentication. The X.509 certificates are issued by trusted third parties called certificate authorities (CA), and the CAs constitute trust federation which guarantees that two different X.509 certificates from different CAs can accomplish authentication against each other. Thus in the GSI based Grid system, if a user would like to access a Grid system, he/she should own a X.509 certificate which is issued by a CA that is trusted by other entities.

In terms of the Virtual Organization issues mentioned earlier, GSI and some other related solutions like the Virtual Organization Management Service (VOMS) ~\cite{AlfieriR05} have partly solved them, to a different extent. The identity delegation and the X.509 proxy certificate initiated by GSI can achieve single sign-on, so that the need to re-enter the user’s pass phrase for authentication can be avoided by creating the proxy certificate.  As the most successful Grid authorization model, VOMS adopts the Attribute Certificate~\cite{RFC3821link} approach. The Attribute Certificate is signed by the VOMS server and then bound to proxy certificates by the client.Attributes include the VO membership data which is associated with the proxy certificate’s identity. Eventually the resources can make access control decisions based on parsing the VO membership attributes from the Attribute Certificate.

The VO is essentially supposed to be able to provide resource sharing capability with resources running under heterogeneous systems/middlewares. This goal can be achieved using Web Service Architecture (WSA). This way, the Web Service technology has been utilized in Grid computing middlewares. The Web Service technology has been adopted by the Globus Toolkit~\cite{GTlink}, gLite~\cite{gLitelink}, as well as the Advanced Resource Connector (ARC)~\cite{ARClink}. In fact, utilizing Web Service technology and providing seamless interoperability with other middlewares is one of the main goals of the new version of ARC middleware.

Unfortunately, the existing Grid security solutions such as GSI and VOMS do not apply to Web Service based applications very well due to the following reasons:
\begin{itemize}
\item The GSS API based confidential communication has not been adopted by Web Service implementations.
\item The Attribute Certificate has not been adopted by any aspects Web Service implementation at all.
\item Web Service implementations commonly use the Web Services Security (WS-Security)~\cite{WSSeclink}, Security Assertion Markup Languages (SAML)~\cite{SAMLlink}, and standard SSL/TLS as well as standard X.509 certificate when considering security, which make them impossible to interoperate with the GSI based Grid middlewares.
\item Finally, identity delegation is a good solution for user or user's representative (acting on behalf of the user) to move/travel from one resource to another. But since the identity delegation has been completely coupled with the GSI confidential communication process, it is not possible for the existing identity delegation solution to be used by Web Service applications.
\end{itemize}

@@

As the key part of the implementation of new ARC middleware, there is a lightweight Web Service container called Hosting Environment Daemon (HED) which provides hosting for various services at application level, as well as a number of modules to support flexible, interoperable, and efficient communication mechanism for building SOAP based Web Services. The whole design of the HED is built around the idea of flexibility and modularity, such that the application service developers can simply concentrate on the application level Web Service implementation by only using the core minimum amount of components. It also simplifies work on the middleware level, e.g. making possible to implement another communication protocol or authentication mechanism by using the core minimum amount of components and external dependencies. A system administrator can easily configure and deploy the middleware and application for different kinds of requirements without having to know much about the implementation.

@@

Dotted line in Figure \ref{fig:HED} shows an alternative path for the information to propagate among MCCs. A service administrator can configure the MCCs according to the interoperability requirements with the other part. For instance, the configuration marked with the dotted line is compatible with the WSE (Web Services Enhancement for .NET) SOAP message mechanism (see WSE's SoapSender and SoapReceiver)~\cite{WSElink}. Another configuration could be SOAP over HTTPg (HTTP over GSI communication protocol) which is needed to interoperate with services like the Storage Resource Manager (SRM)~\cite{AShoshani03} service. This shows the flexibility of HED in terms of protocols support.

\begin{figure}
\includegraphics[width=0.9\columnwidth]{HED.png}
\caption{The example of Host Environment Daemon deployed with A-REX and File services}
\label{fig:HED}
\end{figure}
HED contains a framework for implementing and enforcing authentication and authorization. Each MCC or Service has a common interface for implementing various authentication and authorization functionality. Each functionality can be implemented as a pluggable and configurable component (plug-ins) called SecHandler, which is implemented as a C++ class and provides a method for processing message that travels through MCC or Service. Each MCC or Service is usually configured with two queues of SecHandler -- one for the incoming message and one for the outgoing message respectively. All SecHandler components attached to the queue are executed sequentially. If any of them fails, message processing fails as well. In Figure \ref{fig:HED}, the ``AuthZ'' and ``AuthN'' sub-modules inside MCCs and Services are examples of SecHandler.

As explained above, HED is supposed to act as a Web/Grid Service container, but on the client side, from implementation perspective, there is a similar architecture implemented for processing messages from different protocols and handling security functionality; also there is a specific application programming interface (API) for developers to easily write Web Service client code.

@@

The Shibboleth provides cross-domain single sign-on and attribute-based authorization while preserving user privacy. It is based on the OASIS Security Assertion Markup Language (SAML), specifically, the new version of Shibboleth supports SAML 2.0 specification. For the authentication, the Shibboleth implements SAML2.0 web browser SSO profile, which  defines two functional components, an Identity Provider and a Service Provider. The Identity Provider (IdP) is responsible for creating, maintaining, and managing user identity, while the Service Provider (SP) is responsible for controlling access to services and resources by analyzing the SAML assertions produced and issued by the IdP upon request. In order to discover which home community does a user come from, Shibboleth specifies an optional third component called ``Where Are You From?'' (WAYF) service to aid in the process of IdP discovery. This IdP discovery process is also standardized and defined in SAML 2.0 specification as ``Identity Provider Discovery Profile''.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO.png}
\caption{SAML2.0 SSO profile in ARC}
\label{fig:SAML2SSO}
\end{figure}
Here is the explanation of steps shown in Figure \ref{fig:SAML2SSO}:
\begin{enumerate}
\item The client uses the user agent interface to launch a HTTP request including the IdP name (to which the user belongs) to the service side. The endpoint of the SP (Service Provider) service is the same as that of the other target services, except the last part of the endpoint is ``saml2sp'' which is specific for pointing to the SP Service. Note that we use Identity Provider (IdP) name here to simplify the IdP discovery process in order to avoid the IdP discovery process, because we suppose that the user who would access the target services should better know where is he from initially.
\item The SP Service searches the metadata (we use the same metadata format as defined in the Shibboleth) and gets the location of the single sign-on service (hosted in IdP) and also the location of assertion consuming service (hosted in this SP itself). Then SP Service uses obtained information to compose and issue the SAML authentication request also by utilizing own X.509 certificate (note that in the SAML2.0 SSO profile, X.509 certificates are still needed for IdP and SP) and sends the response back to the user agent.
\item User agent sends the SAML authentication request to the Identity Provider.
\item Identity Provider requires an act of authentication. The authentication mechanism is outside of the SAML 2.0 SSO profile. The Shibboleth IdP implementation uses login handlers for authentication. The current user agent implementation is compatible with the Username/Password login handler of Shibboleth IdP. Through the HTTP protocol, the user agent will feed IdP with the username/password which has been given by the caller of user agent interface.
\item Once the authentication has been succeeded, the IdP issues a SAML response including SAML assertion encrypted by destination SP’s public key, and then this SAML response will be delivered by the user agent to the Service Provider.
\item The SP Service verifies and checks the SAML response, decrypts and stores the SAML assertion into session/connection context. That assertion includes the AuthnStatement and AttributeStatement elements.
\item The WS client launches the Grid/Web Service request via the same connection as the one which is used by user agent to contact SP Service.
\item The Grid/Web Service checks the AuthnStatement stored in the session context to see if the session is still valid. If valid, service handles the  request and returns the response to the WS client. Note that in the current implementation the service requires WS client to reuse same TCP connection as the one used by user agent is step 5, in order to guarantee  validity of the SSO result.
\end{enumerate}
The SP service and other functional service(s) are hosted by the same container, and they use the same X.509 credential for the transport layer security. The client authentication may be switched off, so that client doesn't need to use any X.509 credential. Only the trusted CAs of SP and IdP services need to be configured on the client side to ensure SP and IdP can authentication.

@@

The SLCS client sends request to the SLCS service with the generated X.509 certificate request embedded. Upon reception of the request the SLCS service  composes a distinguished name (DN), andissues a certificate (short lived, 12 hours by default) with the SAML attribute (from the SAML2.0 SSO profile) as the X.509 certificate extension and puts the certificate in to the Web Service response. The SLCS client gets the response and stores the X.509 certificate into local repository along with corresponding private key.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{Delegation_protocol.png}
\caption{Credential delegation protocol}
\label{fig:Delegation_protocol}
\end{figure}
For the delegation process, the delegation WS client functionality (inside each delegation service or the initial client) which is included to access the target delegation service (d1 or d2 in Figure \ref{fig:Delegation}), should use the user's X.509 credential or a delegated credential (on behalf of the user) for authentication and secure communication, as well as delegate one more level of the delegated credential. For the other general  service invocation process, the WS client functionality (inside each service or the initial client) which is included to access the target service (step a1 or a2 in Figure \ref{fig:Delegation}), should also use the user's X.509 credential or a delegated credential for authentication and secure communication.

@@

WS-Security related specifications are implemented as SecHandler in ARC. There are three kinds of WS-Security version 1.1 specifications which are completely or partly implemented: Username Token profile, X.509 Token profile, and SAML Token profile.

For those WS clients on behalf of a user, at any step of service invocation, the WS client can augment the SOAP message with WS-Security token by simply configuring the WS-Security SecHandler into container’s configuration, if the relying party (target service) requires.

@@

The relying party is supposed to trust the issuing authority. When it receives a message from the asserting entity, it will check the SAML assertion based on its predetermined trust relationship with the SAML issuing authority, and check the signature of the SOAP message based on the public key in the SAML assertion without directly trust relationship with attesting entity (holder of this public key).
In terms of verification, since the proxy certificates have been introduced in IETF (RFC 3820)~\cite{RFC3820link}, and OpenSSL has also implemented RFC3820, it is not difficult to use the proxy certificate for the X.509 Token profile to issue an X.509 token.

@@

GSI plug-in for gSOAP~\cite{GAloisio05} adds GSI support for gSOAP Web Service toolkit. The GSI MCC implemented in ARC provides samilar functionality. Moreover, the ARC solution is more flexible because GSI MCC of both service and client sides is pluggable, which means service and client developers of the ARC do not need to modify the service or client implementation in order to add or replace the GSI functionality.

@@

It must be noted that due to the standard specifications utilized (SAML and WS-Security) which are independent from any particular middleware, the WS clients and Web/Grid Services developed within the ARC middleware can interoperate with WS clients and Web/Grid Services developed as part of some other middlewares in a standardized way. ARC also supports interoperability with Grid middlewares that require GSI based communication.

@@

\bibitem{WSElink}
Web Services Enhancements 2.0 Service Pack 2. http://msdn.microsoft.com/en-us/library/

\bibitem{AShoshani03}
A. Shoshani, A. Sim, and J. Gu, Storage Resource Managers: Essential Components for the Grid, Grid Resource Management: State of the Art and Future Trends, Kluwer Publishing, 2003.

@@

\bibitem{RFC3820link}
Internet X.509 Public Key Infrastructure (PKI) Proxy Certificate Profile. ttp://www.ietf.org/rfc/rfc3820.txt

@@

\bibitem{GAloisio05}
G. Aloisio, M. Cafaro, I. Epicoco, D. Lezzi, and R. van Engelen, The GSI plug-in for gSOAP: Enhanced Security, Performance, and Reliability, International Conference on Information Technology: Coding and Computing (ITCC'05), IEEE Press, Volume I, pp. 304-309.

>>>>

\section{Introduction} 
\label{sec:intro}
Grid provides the technology for wide-scale, cross-domain collaboration where users work with resources and share data which are distributed across administrative domains. This kind of collaboration is called Virtual Organization (VO)~\cite{IFoster01}. The resources that contribute to the VO can come from independent administrative domains, and even require different security approaches to access control due to usage of different underlying middlewares. The same heterogeneous situation applies to users who need to access the VO resources, because users holding one kind of a credential may need to utilize multiple resources with different security approaches. To solve the above issues, Grid middlewares should provide the following abilities:
\begin{itemize}
\item The user which participates a VO should be able to access data and resources without being required to constantly do authentication at each resource or service;
\item The VO should be able to easily manage the privileges and memberships of the users;
\item The VO together with its resources should be able to easily manage the authorization, given the condition that the resources are owned by different administrative domains.
\end{itemize}
In the current Grid community, the Grid Security Infrastructure (GSI)~\cite{IFoster98} is the de facto standard for authentication and transport level security. It builds on the Public Key Infrastructure (PKI), an architecture based on the X.509 public key certificate. GSI implements some enhancement (such as certificate delegation) based on the standard SSL/TLS protocol. Mutual authentication is required by GSI, and is the default configuration for GSI based Grid deployment.

In terms of the Virtual Organization issues mentioned earlier, GSI and some other related solutions like the Virtual Organization Management Service (VOMS) ~\cite{AlfieriR05} have partly solved them, to a different extent. The identity delegation and the X.509 proxy certificate initiated by GSI can achieve single sign-on, so that the need to re-enter the user’s pass phrase for authentication can be avoided. VOMS adopts the Attribute Certificate~\cite{RFC3821link} approach. The Attribute Certificate is signed by the VOMS server and then bound to proxy certificates by the client. Attributes include the VO membership data which is associated with the proxy certificate’s identity. Eventually the resources can make access control decisions based on parsing the VO membership attributes from the Attribute Certificate.

The VO is essentially supposed to be able to provide resource sharing capability with resources running under heterogeneous systems/middlewares. This goal can be achieved using Web Service Architecture (WSA). This way, the Web Service technology has been utilized in Grid computing middlewares, including Globus Toolkit~\cite{GTlink}, gLite~\cite{gLitelink}, as well as the Advanced Resource Connector (ARC)~\cite{ARClink}.

Unfortunately, the existing Grid security solutions such as GSI and VOMS do not apply to Web Service based applications very well due to the following reasons:
\begin{itemize}
\item The GSS API based confidential communication and identity delegation has not been adopted by Web Service implementations.
\item The Attribute Certificate has not been adopted by any aspects Web Service implementation at all.
\item Web Service implementations commonly use the Web Services Security (WS-Security)~\cite{WSSeclink}, Security Assertion Markup Languages (SAML)~\cite{SAMLlink}, and standard SSL/TLS as well as standard X.509 certificate when considering security, which make them impossible to interoperate with the GSI based Grid middlewares.
\end{itemize}

@@

As the key part of the implementation of new ARC middleware, there is a lightweight Web Service container called Hosting Environment Daemon (HED) which provides hosting for various services at application level, as well as a number of modules to support flexible, interoperable, and efficient communication mechanism for building SOAP based Web Services. The whole design of the HED is built around the idea of flexibility and modularity.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{HED.png}
\caption{The example of Host Environment Daemon deployed with A-REX and File services}
\label{fig:HED}
\end{figure}
HED contains a framework for implementing and enforcing authentication and authorization. Each MCC or Service has a common interface for implementing various authentication and authorization functionality. Each functionality can be implemented as a pluggable and configurable component (plug-ins) called SecHandler, which is implemented as a C++ class and provides a method for processing message that travels through MCC or Service. Each MCC or Service is usually configured with two queues of SecHandler -- one for the incoming message and one for the outgoing message respectively. In Figure \ref{fig:HED}, the ``AuthZ'' and ``AuthN'' sub-modules inside MCCs and Services are examples of SecHandler.

@@

The Shibboleth provides cross-domain single sign-on and attribute-based authorization while preserving user privacy. It is based on the OASIS Security Assertion Markup Language (SAML) 2.0 specifications. For the authentication, the Shibboleth implements SAML2.0 web browser SSO profile, which  defines two functional components, an Identity Provider and a Service Provider. The Identity Provider (IdP) is responsible for creating, maintaining, and managing user identity, while the Service Provider (SP) is responsible for controlling access to services and resources by analyzing the SAML assertions produced and issued by the IdP upon request.

@@

\begin{figure}
\includegraphics[width=0.7\columnwidth]{SAML2SSO.png}
\caption{SAML2.0 SSO profile in ARC}
\label{fig:SAML2SSO}
\end{figure}
Here is the explanation of steps shown in Figure \ref{fig:SAML2SSO}:
\begin{enumerate}
\item The client uses the user agent interface to launch a HTTP request including the IdP name (to which the user belongs) to the service side. The endpoint of the SP (Service Provider) service is the same as that of the other target services, except the last part of the endpoint is ``saml2sp'' which is specific for pointing to the SP Service. Here Identity Provider (IdP) name is sent in order to avoid the IdP discovery process, because we suppose that the user who would access the target services should better know where is he from initially.
\item The SP Service searches the metadata and gets the location of the single sign-on service (hosted in IdP) and also the location of assertion consuming service (hosted in this SP itself). Then SP Service uses obtained information to compose and issue the SAML authentication request also by utilizing own X.509 certificate (note that in the SAML2.0 SSO profile, X.509 certificates are still needed for IdP and SP) and sends the response back to the user agent.
\item User agent sends the SAML authentication request to the Identity Provider.
\item Identity Provider requires an act of authentication. The authentication mechanism is outside of the SAML 2.0 SSO profile. The Shibboleth IdP implementation uses login handlers for authentication. The current user agent implementation is compatible with the Username/Password login handler of Shibboleth IdP. Through the HTTP protocol, the user agent will feed IdP with the username/password which has been given by the caller of user agent interface.
\item Once the authentication has been succeeded, the IdP issues a SAML response including SAML assertion encrypted by destination SP’s public key, and then this SAML response will be delivered by the user agent to the Service Provider.
\item The SP Service verifies and checks the SAML response, decrypts and stores the SAML assertion into session/connection context. That assertion includes the AuthnStatement and AttributeStatement elements.
\item The WS client launches the Grid/Web Service request via the same connection as the one which is used by user agent to contact SP Service.
\item The Grid/Web Service checks the AuthnStatement stored in the session context to see if the session is still valid. If valid, service handles the  request and returns the response to the WS client. Note that in the current implementation the service requires WS client to reuse same TCP connection as the one used by user agent is step 5, in order to guarantee  validity of the SSO result.
\end{enumerate}
The SP service and other functional service(s) are hosted by the same container, and they use the same X.509 credential for the transport layer security. The client authentication may be switched off, so that client doesn't need to use any X.509 credential. Only the trusted CAs of SP and IdP services need to be configured on the client side to ensure SP and IdP can be authenticated.

@@

The SLCS client sends request to the SLCS service with the generated X.509 certificate request embedded. Upon reception of the request the SLCS service composes a distinguished name (DN), and issues a certificate (short lived, 12 hours by default) with the SAML attribute (from the SAML2.0 SSO profile) as the X.509 certificate extension and puts the certificate in to the Web Service response. The SLCS client gets the response and stores the X.509 certificate into local repository along with corresponding private key.

@@

\begin{figure}
\includegraphics[width=0.7\columnwidth]{Delegation_protocol.png}
\caption{Credential delegation protocol}
\label{fig:Delegation_protocol}
\end{figure}
For the delegation process, the delegation WS client functionality (inside each delegation service or the initial client) which is included to access the target delegation service (d1 or d2 in Figure \ref{fig:Delegation}), should use the user's X.509 credential or a delegated credential (on behalf of the user) for authentication and secure communication, as well as delegate one more level of the delegated credential. For the other general  service invocation process, the WS client functionality (inside each service or the initial client) which is included to access the target service (step a1 or a2 in Figure \ref{fig:Delegation}), should also use the user's X.509 credential or a delegated credential for authentication and secure communication.

@@

WS-Security related specifications are implemented as SecHandler in ARC. There are three kinds of WS-Security version 1.1 specifications which are completely or partly implemented: Username Token profile, X.509 Token profile, and SAML Token profile. For those WS clients on behalf of a user, at any step of service invocation, the WS client can augment the SOAP message with WS-Security token by simply configuring the WS-Security SecHandler into container’s configuration, if the relying party (target service) requires.

@@

The relying party is supposed to trust the issuing authority. When it receives a message from the asserting entity, it will check the SAML assertion based on its predetermined trust relationship with the SAML issuing authority, and check the signature of the SOAP message based on the public key in the SAML assertion without directly trust relationship with attesting entity (holder of this public key).

@@

@@

It must be noted that due to the standard specifications utilized (SAML and WS-Security) which are independent from any particular middleware, the WS clients and Web/Grid Services developed within the ARC middleware can interoperate with WS clients and Web/Grid Services developed as part of some other middlewares in a standardized way.

@@

@@

@@
